\chapter{Proofs by Contradiction}
\label{chapter:indirect-proofs}
\marginurl{%
  Proofs by Contradiction:\\\noindent
  Introduction to Mathematical Reasoning \#3
}{youtu.be/bWP0VYx75DI}
\section{Proving Negative Statements}
The direct method is not very convenient when we need to prove a negation of
some statement.

For example, we may try to prove that $78 n + 102 m = 11$ does not have integer
solutions. It is not clear how to prove it directly since we can not consider
all possible $n$ and $m$. Hence, we need another approach. Let us assume that
such a solution $n, m$ exists. Note that $78 n + 102 m$ is even, but $11$ is
odd. In other words, an odd number is equal to an even number, which is impossible.
Thus, the assumption was false.

Let us consider a more useful example, let us prove that if $p^2$ is even, then
$p$ is also even ($p$ is an integer). Assume the opposite i.e. that $p^2$ is
even but $p$ is not. Let $p = 2b + 1$\footnote{Note that we use here the
statement that an integer $n$ is not even iff it is odd, which, formally
speaking, should be proven.}. Note that $p^2 = (2b + 1)^2 = 2(2b^2 + 2b) + 1$.
Hence, $p^2$ is odd which contradicts to the assumption that $p^2$ is even.

Using this idea we may prove much more complicated results e.g. one may show
that $\sqrt{2}$ is irrational. For the sake of contradiction, let us assume
that it is not true. In other words there are $p$ and $q$ such that
$\sqrt{2} = \frac{p}{q}$ and $\frac{p}{q}$ is an irreducible fraction.

Note that $\sqrt{2} q = p$, so $2q^2 = p^2$. Which implies that $p$ is even
and $4$ divides $p^2$. Therefore $4$ divides $2q^2$ and $q$ is also even. As
a result, we get a contradiction with the assumption that $\frac{p}{q}$ is an
irreducible fraction.

\begin{template}
  \textbf{Template for proving a statement by contradiction.} \\

  Assume, for the sake of contradiction, that \emph{the statement} is false.
  Then \emph{present some argument that leads to a contradiction}. Hence, the
  assumption is false and \emph{the statement} is true.
\end{template}

\begin{exercise}
  Show that $\sqrt{3}$ is irrational.
\end{exercise}

\begin{exercise}
  Let $a$ and $b$ be some integers. Show that $a^2 - 4b - 2 \neq 0$.
\end{exercise}
\begin{solution}
  Assume, for the sake of contradiction, that such $a$ and $b$ exist. Note that
  $a^2 = 4b + 2$, hence, $a$ is even but $a^2$ is not divisible for $4$ which is a
  contradiction.
\end{solution}

\section{Proving Implications by Contradiction}
This method works especially well when we need to prove an implication.
Since the implication $A \implies B$ is false only when $A$ is true but $B$ is
false. Hence, you need to derive a contradiction from the fact that $A$ is true
and $B$ is false.

We have already seen such examples in the previous section, we proved that
$p^2$ is even implies $p$ is even for any integer $p$. Let us consider another
example. Let $a$ and $b$ be reals such that $a > b$. We need to show that
$(ac < bc) \implies c < 0$. So we may assume that $ac < bc$ but $c \ge 0$. By
the multiplicativity of the inequalities we know that if $(a > b)$ and $c > 0$,
then $ac > bc$ which contradicts to $ac < bc$.

A special case of such a proof is when we need to prove the implication
$A \implies B$, assume that $B$ is false and derive that $A$ is false which
contradicts to  $A$ (such proofs are called proofs by contraposition); note
that the previous proof is a proof of this form.

\section{Proof of ``OR'' Statements}
Another important case is when we need to prove that at least one of two
statements is true. For example, let us prove that $ab = 0$ iff $a = 0$ or
$b = 0$. We start from the implication from the right to the left. Since if
$a = 0$, then $ab = 0$ and the same is true for $b = 0$ this implication is
obvious.

The second part of the proof is the proof by contradiction. Assume $ab = 0$,
$a \neq 0$, and $b \neq 0$. Note that $b = \frac{ab}{a} = 0$,
hence $b = 0$ which is a contradiction to the assumption.


\begin{chapterendexercises}
    \exercise[recommended] Prove that if $n^2$ is odd, then $n$ is odd.
    \exercise  In Euclidean (standard) geometry, prove: If two lines share a
        common perpendicular, then the lines are parallel.
        \begin{solution}
          Let us denote by $AB$ the common perpendicular. Assume that the lines
          are not parallel (note that these lines are different) i.e. that there
          is an intersection $C$ of these lines.

          Note that the angles $CAB$ and $CBA$ are right, hence, the angle $ACB$
          is equal to $0$ degrees. So the lines are the same, which is a
          contradiction. 

          Hence, the assumption was incorrect i.e. the lines are parallel.
        \end{solution}
    \exercise[recommended] Let us consider four-lines geometry, it is a theory with
        undefined terms: point, line, is on, and axioms:
        \begin{enumerate}
            \item there exist exactly four lines,
            \item any two distinct lines have exactly one point on both of them, and
            \item each point is on exactly two lines.
        \end{enumerate}

        Show that every line has exactly three points on it.
        \begin{solution}
          Let us assume the opposite; i.e. there is a line $\ell$ with four
          points $p_1, \dots, p_4$ on it. By axiom~3, for each of these points
          $p_i$ there is a line $\ell_i \neq \ell$ such that $p_i$ is on $\ell$.
          Since we have only $4$ lines and $\ell_i \neq \ell$, then there are $i
          \neq j$ such that $\ell_i = \ell_j$. As a result, $p_i$ and $p_j$ are
          on $\ell$ and $\ell_i$, it is contradicting to axiom~2.
        \end{solution}
    \exercise Let us consider group theory, it is a theory with undefined
        terms: group-element and times (if $a$ and $b$ are group elements,
        we denote $a$ times $b$ by $a \cdot b$), and axioms:
        \begin{enumerate}
          \item $(a \cdot b) \cdot c = a \cdot (b \cdot c)$
            for every group-elements $a$, $b$, and $c$;
          \item there is a unique group-element $e$ such that
            $e \cdot a = a = a \cdot e$ for every group-element $a$
            (we say that such an element is the identity element);
          \item for every group-element $a$ there is a group-element $b$
            such that $a \cdot b = e$, where $e$ is the identity element;
          \item for every group-element $a$ there is a group-element $b$
            such that $b \cdot a = e$, where $e$ is the identity element.
        \end{enumerate}

        Let $e$ be the identity element. Show the following statements
        \begin{itemize}
          \item if $b_0 \cdot a = b_1 \cdot a = e$, then $b_0 = b_1$, for every
            group-elements $a$, $b_0$, and $b_1$.
          \item if $a \cdot b_0 = a \cdot b_1 = e$, then $b_0 = b_1$, for every
            group-elements $a$, $b_0$, and $b_1$.
          \item if $a \cdot b_0 = b_1 \cdot a = e$, then $b_0 = b_1$, for every
            group-elements $a$, $b_0$, and $b_1$.
        \end{itemize}
    \exercise Let us consider three-points geometry, it is a theory with
        undefined terms: point, line, is on, and axioms:
        \begin{enumerate}
            \item There exist exactly three points.
            \item Two distinct points are on exactly one line.
            \item Not all the three points are collinear i.e. they do not lay on the
                same line.
            \item Two distinct lines are on at least one point i.e. there is at
              least one point such that it is on both lines.
        \end{enumerate}

        Show that there are exactly three lines.

        \begin{solution}
            Let us denote the points by $p_1$, $p_2$, and $p_3$ (they exist by
            Axiom~1). By Axiom~2, there are lines $l_{1, 2}$,
            $l_{1, 3}$, and $l_{2, 3}$ such that $p_i$ and $p_j$ are on
            $l_{i, j}$ ($i \neq j$).

            Note that the lines $l_{1, 2}$, $l_{1, 3}$, and $l_{2, 3}$ are different.
            Indeed, assume the opposite, i.e., without loss of generality that
            $l_{1, 2} = l_{1, 3}$. Note that $p_1$, $p_2$, and $p_3$ are on $l_{1, 2}$
            which contradicts Axiom~3.

            Let us now prove that there are no other lines. Assume the
            opposite i.e. that there is another line $l$. There is a
            point that is on $l$ and $l_{1, 2}$. Without loss of generality,
            this point is $p_1$. Additionally there is a point $p_i$
            ($i \neq 1$) that is on $l$ and  $l_{2, 3}$. However, it means that
            $p_1$ and $p_i$ are on $l$ which contradicts Axiom~2.
        \end{solution}
    \exercise Show that there are irrational numbers $a$ and $b$ such that
        $a^b$ is rational.
        \begin{solution}
          Assume that $a^b$ is irrational for all irrationals $a$ and $b$. Since
          $\sqrt{2}$ is irrational $\sqrt{2}^{\sqrt{2}}$ is also irrational.
          However, $(\sqrt{2}^{\sqrt{2}})^{\sqrt{2}} = \sqrt{2}^2 = 2$, which
          contradicts to the assumption.

          As a result, there are irrational $a$ and $b$ such that $a^b$ is
          rational.
        \end{solution}
    \exercise[recommended] Show that there does not exist the largest integer.

    \exercise Let us consider Young's geometry, it is a theory with undefined
        terms: point, line, is on, and axioms:
        \begin{enumerate}
            \item there exists at least one line,
            \item every line has exactly three points on it,
            \item not all points are on the same line,
            \item for two distinct points, there exists exactly one line on both of
              them,
            \item if a point does not lie on a given line, then there exists exactly
              one line on that point that does not intersect the given line.
        \end{enumerate}

        Show that for every point, there are exactly four lines on that point.
        \begin{solution}
          First, we prove that for every point, there is a line not on that
          point. Let $p$ be some point. By the first axiom, there is a line
          $\ell$. Assume that $p$ is on $\ell$ (otherwise we proved the
          statement). By axiom 2, there are two other points $p_1$ and $p_2$ on
          this line. By axiom 3, there is a point $q$ not on $\ell$. Finally, by
          axiom 4, there is a line $\ell'$ on $p_1$ and $q$. Note that $\ell'
          \neq \ell$, thus $p$ is not on $\ell'$.
          
          Now we are ready to prove that there are at least four lines. Let $p$
          be a point and $\ell$ be a line such that $p$ is not on $\ell$. By
          axiom 2, there are three points $p_1$, $p_2$, and $p_3$ on $\ell$. By
          axiom 4, there are lines $\ell_1$, $\ell_2$, and $\ell_3$ such that
          $p$ is on all of them and $p_i$ is on $\ell_i$ for $i \in \range{3}$.
          By axiom 5, there is a line $\ell_4$ contain $p$ but $p_1$, $p_2$,
          and $p_3$  are not on $\ell_4$. Thus, there are at least four lines
          through $p$.


          Let us now prove that there is no other line $\ell_5$ such that $p$ is
          on $\ell_5$. By axiom 5, there is a point $p_i$ such that $p_i$ is on
          both $\ell_5$ and $\ell$. But this contradicts to axiom 4, since $p_i$
          is also on $\ell_i$.
        \end{solution}
    \exercise Let us consider five-point geometry, it is a theory with undefined
      terms: point, line, is on, and axioms:
      \begin{enumerate}
        \item there exist exactly five points,
        \item each two distinct points have exactly one line on both of them, and
        \item each line has exactly two points.
      \end{enumerate}

      Show that each point has exactly four lines on it.
      \begin{solution}
        Let us consider some point $p$ and let $p_1$, \dots, $p_4$ be all other
        points (they exist by axiom~1). Note that there are lines $l_1$, \dots,
        $l_4$ such that $l_i$ goes throw $p$ and $p_i$. Note that $l_i \neq l_j$
        for $i \neq j$ by axiom~3 since $p_i$ is not on $l_j$ and there are
        already two points on $l_i$. Thus we proved that any point is on at
        least $4$ lines.

        Let us now prove that there are no other lines. Assume that $p$ is on a
        line $l$ and $l$ is not equal to $l_i$ for all $i$. However, by axiom 2,
        there is a point $p_j$ on $\ell$. Thus $l = l_j$ which is a contradiction.
      \end{solution}
\end{chapterendexercises}
