\chapter{Basic Definitions}
In this part we study what computers can compute and what they cannot compute.
Usually study of this subject starts from a formal definition of an algorithm.
However, we believe that this is unnecessary nowadays because of rise of
computers. One may think about algorithms as programs on some programming
language such as C/C++, Java, Python etc.

An algorithms are taking several natural numbers $x_1$, \dots, $x_n$
as an input and either print another number $y$ as an output or never
terminates. In the first case we say $\Algorithm{A}(x_1, \dots, x_n) = y$
and in the second case we say $\Algorithm{A}(x_1, \dots, x_n)$ never terminates.

\section{Computable Functions}
The first and the most basic definition in the computability theory is the
definition of a computable function.
\begin{definition}
    Let $S \subseteq \N^\ell$ and $f : S \to \N$.
    We say that $f$ is \emph{computable} if there is an algorithm
    $\Algorithm{A}$ such that
    \begin{enumerate}
        \item $\Algorithm{A}(x) = f(x)$ for any $x \in S$ and
        \item $\Algorithm{A}(x)$ never terminates for any $x \notin S$.
    \end{enumerate}
\end{definition}

It is important to note that nonetheless that we say that the algorithms are
said to take and print natural numbers, we could allow algorithm to work with
strings of bits (elements of the set
$\set{0, 1}^* = \bigcup_{n \in \N_0} \set{0, 1}^n$).
Moreover, these two definitions are  equivalent since there is a one-to-one
correspondence between natural numbers and strings:
$x \mapsto 2^n + \sum_{i = 1}^\ell 2^{i - 1} x[i]$, where $x[i]$ denotes the
$i$th symbol of the string $i$ and $n$ is the length of $x$.
It is also clear that using binary strings we may encode all sorts of objects
such as pairs of integers, integers, rational numbers etc. (However, in order
to encode real number we need more complicated definitions and we are not
going to discuss them in here.)

One may show that composition of computable functions is computable.
Moreover, one may prove the following a bit stronger statement.
\begin{theorem}
\label{theorem:composition-computable}
    Let $S \subseteq \N$, and let
    $f_1 : S \to \N$, $f_2 : S \to \N$, and $g : \N^2 \to \N$ be computable
    functions. Then the function $h$ such that $h(x) = g(f_1(x), f_2(x))$
    is also computable.
\end{theorem}

\section{Decidable Sets}
Another important notion is the notion of a decidable set.
\begin{definition}
    We say that a set $S \subseteq \N$ is \emph{decidable} iff there is
    an algorithm $\Algorithm{A}$ such that
    $\Algorithm{A}(x) = 1$ if $x \in S$ and $\Algorithm{A}(x) = 0$ if
    $x \notin S$.
\end{definition}
It is easy to note that a set $S \subseteq \N$ is decidable iff the
characteristic function $\chi_S$ of $S$ is computable.
The function $\chi_S$ is defined as follows:
\[
    \chi_S(x) =
    \begin{cases}
        1 & \text{if } x \in S, \\
        0 & \text{otherwise}.
    \end{cases}
\]
Similarly we may define decidable sets of strings, pairs of integers etc.

Let us illustrate this concept by proving that the set
$U_e = \set[q > e]{q \in \Q}$. It is known that
\begin{enumerate}
    \item $(1 + \frac{1}{n})^n < e$ and
        $\lim\limits_{n \to \infty} (1 + \frac{1}{n})^n = e$ and
    \item $(1 + \frac{1}{n})^{n + 1} > e$ and
        $\lim\limits_{n \to \infty} (1 + \frac{1}{n})^{n + 1} = e$.
\end{enumerate}
Hence, in order to check whether $q \in U_e$ or not, it is enough to either find
$n$ such that $(1 + \frac{1}{n})^{n + 1} < q$ or $(1 + \frac{1}{n})^n > q$.
In order to do this we can check all positive integers one after another
and at some point one of the inequalities became true.

\begin{exercise}
  Let $k \in \N$. Show that $[k]$ is decidable.
\end{exercise}

However, sometimes it is possible to show that some set is decidable without
presenting the algorithm explicitly.
For example, consider the $S \subseteq \N$ such that $n \in S$ iff base $10$
expansion of $\pi$ has $n$ consecutive $9$s. It is possible to show that $S$ is
decidable. Indeed, it is easy to see that either $S = \N$ or $S = [k]$ for some
$k \in \N$, however, in both cases the set is decidable.


\section{Enumeratable Sets}
The algorithm constructed in the proof of decidability of $U_e$ consisted of
two important parts: first one allowed to show that $q$ is defenetely not in the
set $U_e$ and the second part allowed to show that $q$ is defenetely in the set.

This observation leads to the following definition.
\begin{definition}
  We say that a set $S \subseteq \N$ is \emph{(computably) enumeratable},
  we also say that it is \emph{recursively enumerable} and \emph{semidecidable},
  iff there is an algorithm $\Algorithm{A}$ such that
  \begin{enumerate}
    \item $\Algorithm{A}(x) = 1$ for any $x \in S$ and
    \item either $\Algorithm{A}(x) = 0$ or $\Algorithm{A}$ never terminates for
      any $x \notin S$.
  \end{enumerate}
\end{definition}

We can easily show that both sets $L_e = \set[q < e]{q \in \Q}$ and
$U_e = \set[q > e]{q \in \Q}$ are enumeratable. Indeed, we can try all possible
$n \in n$ until $q < (1 + \frac{1}{n})^n$ if we find such $n$, we know that
$q \in L_e$. Similarly, we can try all possible $n$ until
$q > (1 + \frac{1}{n})^{n + 1}$ and if we find such $n$, then $q \in U_e$.
The following allows us to show that the fact that $L_e$ is decidable is not
a coincedence.
\begin{theorem}
  Let $S \subseteq \N$. If $S$ is decidable, then $S$ is enumeratable.
  Moreover, if $S$ and $\N \setminus S$ are enumeratable, then $S$ is
  decidable.
\end{theorem}

The given definition of enumeratable set does not explain the name. However,
there is an alternative definition that explains it.
\begin{theorem}
  Let $S \subseteq \N$. The set $S$ is decidable iff there is an algorithm
  $\Algorithm{A}$ such that
  \begin{enumerate}
    \item $\Algorithm{A}(n)$ terminates for any $n \in \N$ and
    \item $\set[n \in  \N]{\Algorithm{A}(n)} = S$.
  \end{enumerate}
\end{theorem}
Informally, the algorithm $\Algorithm{A}$, from this theorem, enumerates
all the elements of $S$.

One may also esteblish a connection between computable functions and
enumeratable sets.
\begin{theorem}
  Let $S \subseteq \N$.
  \begin{enumerate}
    \item The set $S$ is enumeratable iff there is a computable function
      $f : S \to \N$.
    \item The set $S$ is enumeratable iff there is a computable function
      $f : \N \to S$.
  \end{enumerate}
\end{theorem}
